What we {\sf eval} in the shadows

### 2021-03-31 analysis/environments.Rmd

  E_raw %>% select(
    run_package = package, #      The package which was run to trigger the eval
    ev_variant = eval_function, # Which eval variant was called
    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
    src_ref = eval_call_srcref, # Source ref for the eval call site
    file, #                       Script file name (name of this run)
    ev_package = eval_source, #   Package of the call site
    type = expr_resolved_type_tag, #   Argument type
    ev_expr = expr_expression,
    n_op = interp_eval,
    envir = envir_expression,
    envir_type,
    caller_expression,
    environment_class
  ) -> E_envir

@@

```{r envir_most_common}
frequent_envir <- E_envir %>% group_by(src_ref) %>% 
  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "))

stopifnot(frequent_envir$n_envir <= 1)
```

# `envir` with Rcpp

```{r envir_rcpp}
eval_calls_rcpp <- eval_calls_explicit %>% filter(str_detect(expr_expression, "C\\+\\+Constructor|C\\+\\+Field|C\\+\\+OverloadedMethods"))

nb_rcpp <- count(eval_calls_rcpp, wt = duplicates)
nb_rcpp_sites <- nb_eval_call_sites(eval_calls_rcpp)

eval_calls_explicit <- eval_calls_explicit %>% anti_join(eval_calls_rcpp)
```

@@

nb_env_list <- count(eval_calls_env_list, wt = duplicates)
nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)
```

@@

```{r envir_arg_caller}
eval_calls_arg_caller <- eval_calls_explicit %>% filter(!is.na(envir_from_arg))

@@

```{r}
eval_calls_arg_caller %>%
  count(envir_from_arg, wt = duplicates, sort = TRUE) %>%
  knitr::kable()
```

@@

```{r envir_absolute}
eval_calls_absolute <- eval_calls_explicit %>%
  anti_join(eval_calls_env_list) %>%
  anti_join(eval_calls_arg_caller)

nb_absolute <- count(eval_calls_absolute, wt = duplicates)
nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)
```

@@

```{r}
eval_calls_absolute <- eval_calls_absolute %>%
  group_by(eval_call_srcref) %>%
  mutate(envir_distinct = n_distinct(environment_class))

eval_calls_absolute %>%
  ungroup() %>%
  count(envir_distinct, sort = TRUE)
```

We can also see a few examples:

```{r}
eval_calls_absolute %>%
  filter(envir_distinct == 2) %>%
  select(ev_expr, environment_class, envir_distinct)
```

@@

@@

env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(eval_source, wt = duplicates, sort = TRUE)
```

@@

nb_integer <- count(eval_calls_integer, wt = duplicates)
nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)
```

@@

nb_null <- count(eval_calls_null, wt = duplicates)
nb_null_sites <- nb_eval_call_sites(eval_calls_null)
```

@@

```{r summary}
summary <- tribble(
  ~category, ~calls, ~percent_calls, ~callsites, ~percent_callsites,
  "Default", nb_default$n, ratio(nb_default$n, nb_eval_calls$n), nb_default_sites, ratio(nb_default_sites, nb_call_sites),
  "Rcpp", nb_rcpp$n, ratio(nb_rcpp$n, nb_eval_calls$n), nb_rcpp_sites, ratio(nb_rcpp_sites, nb_call_sites),
  "List or dataframe", nb_env_list$n, ratio(nb_env_list$n, nb_eval_calls$n), nb_env_list_sites, ratio(nb_env_list_sites, nb_call_sites),
  "Argument of caller", nb_arg_caller$n, ratio(nb_arg_caller$n, nb_eval_calls$n), nb_arg_caller_sites, ratio(nb_arg_caller_sites, nb_call_sites),
  "Absolute", nb_absolute$n, ratio(nb_absolute$n, nb_eval_calls$n), nb_absolute_sites, ratio(nb_absolute_sites, nb_call_sites),
  "Integer", nb_integer$n, ratio(nb_integer$n, nb_eval_calls$n), nb_integer_sites, ratio(nb_integer_sites, nb_call_sites),
  "Null", nb_null$n, ratio(nb_null$n, nb_eval_calls$n), nb_null_sites, ratio(nb_null_sites, nb_call_sites),
)

>>>>

  E_raw %>% select(
    run_package = package, #      The package which was run to trigger the eval
    ev_variant = eval_function, # Which eval variant was called
    duplicates = nb_ev_calls, #   Number of identical eval calls (weight)
    src_ref = eval_call_srcref, # Source ref for the eval call site
    file, #                       Script file name (name of this run)
    ev_package = eval_source, #   Package of the call site
    type = expr_resolved_type_tag, #   Argument type
    ev_call = eval_call_expression,
    ev_expr = expr_expression,
    n_op = interp_eval,
    envir = envir_expression,
    envir_type,
    caller_expression,
    environment_class
  ) -> E_envir

@@

```{r envir_most_common}
frequent_envir <- E_envir %>% group_by(src_ref) %>% 
  summarize(n_envir = n_distinct(envir), envir = paste0(unique(envir), collapse = " ; "), envir_type = paste0(unique(envir_type), collapse = " ; "))

#stopifnot(frequent_envir$n_envir <= 1)
```

```{r}
freq_envir <- frequent_envir %>% count(envir, sort = TRUE) %>% 
  mutate(percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent))

nb_parent.frame_sites <- freq_envir %>% filter(envir == "parent.frame()") %>% .$n
nb_data_sites <- freq_envir %>% filter(envir == "data") %>% .$n

overview_table(
  r("nb parent france sites", nb_parent.frame_sites),
  r("nb parent france site percent", ratio(nb_parent.frame_sites, nb_call_sites)),
  r("nb data sites", nb_data_sites),
  r("nb data sites", ratio(nb_data_sites, nb_call_sites))
)
```

```{r}
freq_envir %>% filter(cum_percent <= 90) %>%
  ggplot() +
  geom_col(aes(x = fct_reorder(envir, percent), y = percent)) + 
  coord_flip()

ggsave_prefix("envir_expressions.pdf")
```

Let's classify the expressions, between calls and variables:

```{r}
envir_var <- frequent_envir %>% filter(envir != "NA", str_detect(envir, "^(?:\\.[:alpha]|[:alpha:])[[:alpha:][:digit:]_\\.]*$"))

nb_envir_var_sites <- nrow(envir_var)

overview_table(
  r("envir variable sites", nb_envir_var_sites),
  r("envir variable site percent", ratio(nb_envir_var_sites, nb_explicit_sites))
)
```

What about we classify them by type?

```{r}
frequent_envir %>% count(envir_type, sort = TRUE) %>% datatable()
```

@@

nb_env_list <- count(eval_calls_env_list, wt = duplicates)
nb_env_list_sites <- nb_eval_call_sites(eval_calls_env_list)

overview_table(
  r("nb env list", nb_env_list),
  r("nb env list sites", nb_env_list_sites),
  r("nb env list site percent", ratio(nb_env_list_sites, nb_call_sites))
)
```

@@

```{r envir_arg_caller, eval = FALSE}
eval_calls_arg_caller <- eval_calls_explicit %>% filter(!is.na(envir_from_arg))

@@

```{r eval = FALSE}
eval_calls_arg_caller %>%
  count(envir_from_arg, wt = duplicates, sort = TRUE) %>%
  knitr::kable()
```

@@

```{r envir_absolute}
eval_calls_absolute <- eval_calls_explicit %>%
  anti_join(eval_calls_env_list) #%>%
  #anti_join(eval_calls_arg_caller)

nb_absolute <- count(eval_calls_absolute, wt = duplicates)
nb_absolute_sites <- nb_eval_call_sites(eval_calls_absolute)

overview_table(
  r("nb absolute envir", nb_absolute$n),
  r("nb absolute sites", nb_absolute_sites),
  r("nb absolute site percent", ratio(nb_absolute_sites, nb_eval_calls))
)
```

@@

```{r}
absolute_diversity <- eval_calls_absolute %>%
  group_by(src_ref) %>%
  summarize(envir_distinct = n_distinct(environment_class), envir = paste0(environment_class, collapse = " ; "))

absolute_diversity %>%
  count(envir_distinct, sort = TRUE)
```

@@

```{r}
env_calls %>% group_by(new_env) %>% summarize(n = n_distinct(src_ref), percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent)) %>% arrange(desc(n))

env_calls %>% group_by(hierarchy) %>% summarize(n = n_distinct(src_ref), percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent)) %>% arrange(desc(n))

@@

env_calls %>% filter(new_env == 4, specific_env == "global") %>% count(ev_package, wt = duplicates, sort = TRUE)
```

# Specific environments

```{r}
specific_env <- env_calls %>% group_by(specific_env) %>% summarize(n = n_distinct(src_ref), n_packages = n_distinct(ev_package), percent = round(100 * n / nb_call_sites, 2), cum_percent = cumsum(percent)) %>% arrange(desc(n))

nb_global_env_sites <- specific_env %>% filter(specific_env == "global") %>% .$n
nb_callee_env_sites <- specific_env %>% filter(specific_env == "callee") %>% .$n
nb_empty_env_sites <- specific_env %>% filter(specific_env == "empty") %>% .$n
nb_base_env_sites <- specific_env %>% filter(specific_env == "base") %>% .$n

overview_table(
  r("nb global env sites", nb_global_env_sites),
  r("nb global env site percent", ratio(nb_global_env_sites, nb_call_sites)),
  r("nb callee env sites", nb_callee_env_sites),
  r("nb callee env site percent", ratio(nb_callee_env_sites, nb_call_sites)),
  r("nb empty env sites", nb_empty_env_sites),
  r("nb empty env site percent", ratio(nb_empty_env_sites, nb_call_sites)),
  r("nb base env sites", nb_base_env_sites),
  r("nb base env site percent", ratio(nb_base_env_sites, nb_call_sites))
)
```

## Package environments

Our tracer leaks here: 3 srcref are seeing `package:evil`, from `test_that`, `R.oo` and `envnames`.

What are the ones looking into packages?

```{r}
package_env <- E_envir %>% filter(str_detect(environment_class, fixed("package:"))) %>% count(ev_package, sort = TRUE)

nb_package_env_sites <- E_envir %>% filter(str_detect(environment_class, fixed("package:"))) %>% pull(src_ref) %>% n_distinct()

overview_table(
  r("nb package env packages", nrow(package_env)),
  r("nb package env sites", nb_package_env_sites),
  r("package env site percent", ratio(nb_package_env_sites, nb_call_sites))
)
```

They are the following:

- _R.oo_: OOP system
- _USSCXenaTools_: uses `eval` to create _hidden_ function inside its own namespace (by pasting `".p_"` in front of the function name)
- _unitizer_: unit test framework that requires to precise the test environment
- _envnames_: seach env by names, give a name for each environments. Traverse all the call sites with `sys.calls`
- _testthat_: unit test framework

## Weird packages

There are 3 weird _package_ environments: `.R_Cache`, `.R_Test`, `newdir...`

They all seem to implement a kind of caching mechanism.

They build the environment name with:

```{r, eval = FALSE}
env <- attach(NULL, pos, basename(directory))
```

@@

nb_integer <- count(eval_calls_integer, wt = duplicates)$n
nb_integer_sites <- nb_eval_call_sites(eval_calls_integer)

overview_table(
  r("nb integer envir", nb_integer),
  r("nb integer envir sites", nb_integer_sites),
  r("nb integer envir site percent", ratio(nb_integer_sites, nb_call_sites))
)
```

@@

nb_null <- count(eval_calls_null, wt = duplicates)$n
nb_null_sites <- nb_eval_call_sites(eval_calls_null)

overview_table(
  r("nb null envir", nb_null),
  r("nb null envir sites", nb_null_sites),
  r("nb null envir site percent", ratio(nb_null_sites, nb_eval_calls))
)
```

@@

```{r summary, eval = FALSE}
summary <- tribble(
  ~category, ~calls, ~percent_calls, ~callsites, ~percent_callsites,
  "Default", nb_default$n, ratio(nb_default$n, nb_eval_calls$n), nb_default_sites, ratio(nb_default_sites, nb_call_sites),
  "Rcpp", nb_rcpp$n, ratio(nb_rcpp$n, nb_eval_calls$n), nb_rcpp_sites, ratio(nb_rcpp_sites, nb_call_sites),
  "List or dataframe", nb_env_list$n, ratio(nb_env_list$n, nb_eval_calls$n), nb_env_list_sites, ratio(nb_env_list_sites, nb_call_sites),
  "Argument of caller", nb_arg_caller$n, ratio(nb_arg_caller$n, nb_eval_calls$n), nb_arg_caller_sites, ratio(nb_arg_caller_sites, nb_call_sites),
  "Absolute", nb_absolute$n, ratio(nb_absolute$n, nb_eval_calls$n), nb_absolute_sites, ratio(nb_absolute_sites, nb_call_sites),
  "Integer", nb_integer$n, ratio(nb_integer$n, nb_eval_calls$n), nb_integer_sites, ratio(nb_integer_sites, nb_call_sites),
  "Null", nb_null$n, ratio(nb_null$n, nb_eval_calls$n), nb_null_sites, ratio(nb_null_sites, nb_call_sites),
)

### 2021-03-31 paper/main.tex

\subsubsection{The expression passed to \eval.}

>>>>

\subsubsection{The expression passed to \eval.}
\paragraph{Touching package environments.} It refers to the \emph{namespace} of a package. A call to \c{library(packageName)}, which load libraries in R, for instance, will populate environment \c{package:packageName}. It is a \emph{base} function though, so we do not see it in the CRAN dataset. However, R still makes it possible to access any package environments and \packageNbPackageEnvSites call sites, \ie \packagePackageEnvSitePercent access a package environment, from \packageNbPackageEnvPackages different packages. Those packages perform introspection, to represent and search any environment by name, in package \emph{envnames}, to containerize tests, as with \emph{testthat} and \emph{unitizer}, to create a custom OOP system with \emph{R.oo}, or to create poor man's hidden functions in \emph{USSCXenaTools}.

It is also used to implement a custom caching mechanism in packages \emph{g.data} and \emph{SOAR}.
